---
title: "eppoFindeR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{eppoFindeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

**eppoFindeR** provides an R interface to the public APIs of the **European and Mediterranean Plant Protection Organization (EPPO)** database. The package facilitates access to a wide range of pest-related information collected and maintained by EPPO, allowing users to query, retrieve, and process this data directly from R.

The package is intended for researchers, analysts, and practitioners in plant protection who require convenient programmatic access to EPPO data.

## Installation

### From CRAN

```{r eval=FALSE}
install.packages("eppoFindeR")
```

### Development version (from GitHub)

To install the latest development version:

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("openefsa/eppoFindeR")
```

## Requirements

An active internet connection is required, as the package communicates with EPPO’s online services to fetch and process data.

## Working with API keys

The *eppoFindeR* package requires your unique API token to function properly. You can provide this token in one of two ways:

1.  By setting it in the `.Renviron` file.\
2.  By including it manually in each API request.

### Setting the API key via `.Renviron`

The `.Renviron` file is used to define environment variables that R loads automatically at the start of each session. This approach is particularly convenient for sensitive information like API keys, as it allows you to use them in any R script or function without hardcoding them.

Place the `.Renviron` file in your home directory (for example, `C:/Users/username/Documents/.Renviron` on Windows or `~/.Renviron` on Unix-like systems). You can create or edit this file with any plain text editor.

Add your EPPO API key in the following format:

`EPPO_API_KEY=<your_eppo_api_key>`

After saving the file, R will automatically read the token on startup.

### Setting the API key manually for each request

Alternatively, you can provide the API key directly in the `apiKey` argument of functions that require it. This is useful if you want to use different tokens for different calls, or if you prefer not to store the token globally. For example:

```{r apiKeyExample, eval=FALSE}
taxon(c("BEMITA"), apiKey = "<your_eppo_api_key>")
```

## Basic usage

The main purpose of *eppoFindeR* is to query the EPPO database for specific EPPO codes and retrieve relevant information across various endpoints. For each endpoint, you can either:

1.  Query all available services to obtain more comprehensive data.\
2.  Query one or more services to obtain specific data.

For plant and pest species, the basic information returned includes scientific names, synonyms, common names in different languages, and taxonomic classification. For pests of regulatory interest, you can also retrieve more detailed information, such as host plants and quarantine categorization.

Below are examples demonstrating how to use the functions in this package. First, load the *eppoFindeR* package:

```{r loadLibrary, eval=FALSE}
library(eppoFindeR)
```

To explore the arguments and usage of a specific function, you can run:

```{r eval=FALSE}
help("<function_name>")
```

This will show the full documentation for the function, including its arguments, return values, and usage examples.

For example, if you are working with the `taxon()` function, you can check its documentation with:

```{r eval=FALSE}
help("taxon")
```

## Querying a specific category

The *eppoFindeR* package allows you to query all categories available in version 2.0 of the EPPO APIs: General, Taxons, Taxon, Country, Tools, Reporting Services, and References.

Each category has a corresponding function in the package with the same name in *lowerCamelCase* format: general(), taxons(), taxon(), country(), tools(), reportingService(), and reportings(). By default, these functions return all data available under the selected category, but you can customize the query by specifying the desired services.

For example, to query all services of the Taxon category for the EPPO code "BEMITA", you can use the following code:

```{r taxonAllServices, eval=FALSE}
taxon(c("BEMITA"))
```

Some services require certain mandatory parameters, which must be provided when making the request. For example:

```{r toolsName2Codes, eval=FALSE}
tools(
  c("name2codes"), # The service to query (/tools/name2codes).
  params = list(
    name2codes = list(
      name = "Bemisia tabaci" # The required parameter.
    )
  )
)
```

## Querying a specific service

To query a specific service within an endpoint, eppoFindeR allows you to specify it directly in the function call. For example, to access only the `/taxons/taxon/categorization` service for the EPPO code "BEMITA", you can use the `taxon()` function as follows:

```{r queryService, eval=FALSE}
taxon(c("BEMITA"), c("categorization"))
```

### Expected output

```{r queryServiceOutput, eval=FALSE}
$categorization
# A tibble: 20 × 12
   queried_eppo_code continent_id continent_name country_iso country_name    qlist qlist_label year_add
   <chr>                    <int> <chr>          <chr>       <chr>           <chr> <chr>          <int>
 1 BEMITA                       4 America        CL          Chile           1     A1 list         2019
 2 BEMITA                       2 Asia           BH          Bahrain         2     A2 list         2003
 3 BEMITA                       2 Asia           KZ          Kazakhstan      1     A1 list         2017
 4 BEMITA                       1 Europe         AZ          Azerbaijan      1     A1 list         2024
 5 BEMITA                       1 Europe         BY          Belarus         X     Quarantine…     1994
 6 BEMITA                       1 Europe         GE          Georgia         1     A1 list         2018
 7 BEMITA                       1 Europe         MD          Moldova, Repub… X     Quarantine…     2017
 8 BEMITA                       1 Europe         NO          Norway          X     Quarantine…     2012
 9 BEMITA                       1 Europe         RU          Russian Federa… 2     A2 list         2014
10 BEMITA                       1 Europe         RS          Serbia          1     A1 list         2015
11 BEMITA                       1 Europe         CH          Switzerland     1     A1 list         2019
12 BEMITA                       1 Europe         TR          Türkiye         2     A2 list         2016
13 BEMITA                       1 Europe         UA          Ukraine         1     A1 list         2019
14 BEMITA                       1 Europe         GB          United Kingdom  1     A1 list         2020
15 BEMITA                       5 Oceania        NZ          New Zealand     X     Quarantine…     2000
16 BEMITA                       6 RPPO/EU        9M          EAEU            2     A2 list         2016
17 BEMITA                       6 RPPO/EU        9A          EPPO            2     A2 list         1989
18 BEMITA                       6 RPPO/EU        9L          EU              P     PZ Quarant…     2019
19 BEMITA                       6 RPPO/EU        9L          EU              N     A1 Quarant…     2019
20 BEMITA                       6 RPPO/EU        9H          OIRSA           2     A2 list         1992
# ℹ 4 more variables: year_delete <lgl>, year_transient <lgl>, queried_url <chr>, queried_on <dttm>
```

All function calls return a list of dataframes, each containing the information from the requested services.

## Querying all available services

To fetch data from all available services within a specific endpoint, simply use the corresponding function without modifying the services parameter. Each function will return a list of dataframes, with each dataframe containing information from one service.

```{r queryWhole, eval=FALSE}
data <- taxon(c("BEMITA"))
```

### Expected output

```{r queryAllOutput, eval=FALSE}
# Print the names of the services in the result.
names(data)
 [1] "overview"           "infos"              "names"              "taxonomy"
 [5] "categorization"     "kingdom"            "hosts"              "pests"           
 [9] "vectors"            "vectorof"           "bca"                "bcaof"           
[13] "photos"             "reporting_articles" "documents"          "standards"       
[17] "distribution"      

# Print the data from a specific service.
data$overview
# A tibble: 1 × 11
  queried_eppo_code eppocode datecreate         lastupdate infos replacedby prefname is_active datatype
  <chr>             <chr>    <chr>              <chr>      <lgl> <lgl>      <chr>    <lgl>     <chr>   
1 BEMITA            BEMITA   2002-10-28 00:00:… 2002-10-2… NA    NA         Bemisia… TRUE      GAI     
# ℹ 2 more variables: queried_url <chr>, queried_on <dttm>
```

Using `names(data)` will display the names of the queried services. You can then access a specific service to view the data associated with it.

## Data wrangling

The *eppoFindeR* package provides a convenient function, `uniformTaxonomy()`, to create a complete and standardized taxonomy dataframe. This function takes the taxonomy data returned by the taxonomy service and ensures a uniform structure, including all expected taxonomic ranks, even if some ranks are missing in the original result.

For example, assume you have retrieved taxonomy data for the EPPO code "BEMITA" using the `taxon()` function:

```{r eval=FALSE}
data <- taxon(c("BEMITA"), c("taxonomy"))
```

You can then generate a uniform taxonomy table with all ranks using:

```{r uniformTaxonomy, eval=FALSE}
taxonomy <- uniformTaxonomy(data$taxonomy)
```

### Expected output

```{r uniformTaxonomyOutput, eval=FALSE}
# A tibble: 11 × 6
   type      queried_eppo_code eppocode prefname       queried_url                  queried_on         
   <chr>     <chr>             <chr>    <chr>          <chr>                        <dttm>             
 1 Kingdom   BEMITA            1ANIMK   Animalia       https://api.eppo.int/gd/v2/… 2025-11-18 14:29:50
 2 Phylum    BEMITA            1ARTHP   Arthropoda     https://api.eppo.int/gd/v2/… 2025-11-18 14:29:50
 3 Subphylum BEMITA            1HEXAQ   Hexapoda       https://api.eppo.int/gd/v2/… 2025-11-18 14:29:50
 4 Class     BEMITA            1INSEC   Insecta        https://api.eppo.int/gd/v2/… 2025-11-18 14:29:50
 5 Subclass  NA                NA       NA             NA                           NA                 
 6 Order     BEMITA            1HEMIO   Hemiptera      https://api.eppo.int/gd/v2/… 2025-11-18 14:29:50
 7 Suborder  BEMITA            1STERR   Sternorrhyncha https://api.eppo.int/gd/v2/… 2025-11-18 14:29:50
 8 Family    BEMITA            1ALEYF   Aleyrodidae    https://api.eppo.int/gd/v2/… 2025-11-18 14:29:50
 9 Subfamily NA                NA       NA             NA                           NA                 
10 Genus     BEMITA            1BEMIG   Bemisia        https://api.eppo.int/gd/v2/… 2025-11-18 14:29:50
11 Species   BEMITA            BEMITA   Bemisia tabaci https://api.eppo.int/gd/v2/… 2025-11-18 14:29:50
```

## Data retrieval for multiple EPPO or ISO codes

The `taxon()` and `country()` functions allow you to retrieve also a complete data dump for multiple EPPO or ISO codes by querying all available EPPO services for each of them.

Here is an example showing how to use the function to fetch all available information for the EPPO codes "BEMITA" and "GOSHI":

```{r multipleEPPOCodes, eval=FALSE}
data <- taxon(c("BEMITA", "GOSHI"))
```

### Expected output

```{r multipleEPPOCodesOutput, eval=FALSE}
# Print the names of the services in the result.
names(data)
 [1] "overview"           "infos"              "names"              "taxonomy"
 [5] "categorization"     "kingdom"            "hosts"              "pests"           
 [9] "vectors"            "vectorof"           "bca"                "bcaof"           
[13] "photos"             "reporting_articles" "documents"          "standards"       
[17] "distribution"      

# Print the data from a specific service.
data$overview
# A tibble: 2 × 11
  queried_eppo_code eppocode datecreate         lastupdate infos replacedby prefname is_active datatype
  <chr>             <chr>    <chr>              <chr>      <chr> <lgl>      <chr>    <lgl>     <chr>   
1 BEMITA            BEMITA   2002-10-28 00:00:… 2002-10-2… NA    NA         Bemisia… TRUE      GAI     
2 GOSHI             GOSHI    2004-04-21 00:00:… 2004-04-2… <p>M… NA         Gossypi… TRUE      PFL     
# ℹ 2 more variables: queried_url <chr>, queried_on <dttm>
```

Using `names(data)` will display the names of the queried services. You can then access a specific service to view the data associated with it.
