---
title: "eppoFindeR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{eppoFindeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
library(eppoFindeR)
```

# Overview
Welcome to the _eppoFindeR_ vignette! This package is a wrapper around the public APIs of the European and Mediterranean Plant Protection Organization (EPPO) database. It offers a convenient and efficient way to access a wide range of information on plant pests, diseases, and species using R, making it an invaluable tool for researchers and practitioners in plant protection.

## Prerequisites
To use the package, ensure you have a stable internet connection, as it will not work otherwise. The package communicates with the EPPO database online services to fetch and manipulate data, so a reliable connection is essential.

## Installation
You can install the eppoFindeR package from CRAN with the following command:

```{r instCRAN}
# install.packages("eppoFindeR")
```

Or, if you are installing from Azure DevOps:

```{r instDevOps}
# install.packages("devtools")
# devtools::install_github("yourusername/eppoFindeR")
```

# Basic Usage
The primary functionality of _eppoFindeR_ is to retrieve data about EPPO codes. Users have two main options: 1) only fetch data from a specific service, or 2) fetch all data from the database (i.e. query all available services).
 
For example, to query the categorization service for the "BEMITA" EPPO code, you can use the queryTheEPPOForService() function.

```{r queryServ}
library(eppoFindeR)

result <- queryTheEPPOForService("BEMITA", service = "categorization")
print(result)
```

Alternatively, you can query the entire EPPO database, which means performing a search across all the available services. The result will be a list of data frames, each corresponding to a different service (e.g. pests, plants, diseases etc.). 
 
```{r queryWhole}
result_list <- queryTheWholeEPPO("BEMITA")

# Print the names of the services in the result
print(names(result_list))

# Print the data from a specific service
print(result_list$categorization)
```

The eppoFindeR package also includes a data wrangling function that helps combine the data fetched from the taxonomy and rank services, thus allowing for a more comprehensive and integrated data retrieval.

Assume 'taxonomy' and 'kingdom' are two dataframes obtained from querying the corresponding EPPO services:

```{r rankExample}
taxonomy <- queryTheEPPOForService("BEMITA", service = "taxonomy")
kingdom <- queryTheEPPOForService("BEMITA", service = "kingdom")

# Wrangle data to combine taxonomy and rank
wrangled_data <- taxonomyRanked(taxonomy, kingdom)
print(wrangled_data)
```

The package also implements a data dump function that queries all the available EPPO services, given a vector of input EPPO codes. This ensures thorough data retrieval for multiple EPPO codes.

```{r backUpExample}
# Perform a backup search across all services for a pool of EPPO codes
backup_result <- backUpEPPO(c("BEMITA", "LEUCSC"))
print(backup_result)
```

